Normalization by Evaluation for Sized Dependent Types

### 2017-02-21 nbe/icfp17.tex
[ ICFP 17 ] Proved subsumption.

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (a + 1) \qquad
      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, (\zero{a}) \qquad
      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat a \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

@@

\fbox{$\Gamma \der T \leq T'$} (implies $\resurrect \Gamma \der T,T'$)
\begin{gather*}
  \ru{\der \Gamma \qquad \ell {\leq} \ell'
    }{\Gamma \der \Set[\ell] \leq \Set[\ell']}
\qquad
  \ru{\Gamma \der a \leq b : \Size
    }{\Gamma \der \Nat a \leq \Nat b}
\qquad
  \ru{\Gamma \der T = T'
    }{\Gamma \der T \leq T'}
\\[2ex]
  \ru{\Gamma \der U' \leq U \qquad
      \cext \Gamma {U'} \der T \leq T'
    }{\Gamma \der \piT U T \leq \piT {U'} {T'}}
\qquad
  \ru{\Gamma \der T_1 \leq T_2 \qquad
      \Gamma \der T_2 \leq T_3
    }{\Gamma \der T_1 \leq T_3}
\qquad
\end{gather*}

@@

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \alpha \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{(\zero \alpha) @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc \alpha t) @ \casett \evalsto w}
\\[2ex]
  \ru{t\,\ann \alpha\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero \alpha, \suc \alpha u \}}
\end{gather*}
The rules added for NbE deal with elimination of delayed reflection:
\begin{gather*}
  \ru{T' \evalsto \piT U T
    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
\qquad
  \ru{T' \evalsto \piT \Size T
    }{(\up {T'} n) @ \alpha \evalsto \up{T[\alpha]} (n \, \alpha)}
\qquad
  \ru{T' \evalsto \forallT T
    }{(\up {T'} n) @ \ann \alpha \evalsto \up{T[\alpha]} (n \ann \alpha)}
%   \ru{T \evalsto \piT A B \qquad
%       \up{B[u]} (n \, \down A u) \evalsto w
%     }{(\up T n) @ u \evalsto w}
% \qquad
%   \ru{T \evalsto \piT \Size B \qquad
%       \up{B[u]} (n \, \alpha) \evalsto w
%     }{(\up T n) @ \alpha \evalsto w}
% \qquad
%   \ru{T \evalsto \forallT B \qquad
%       \up{B[\alpha]} (n \ann \alpha) \evalsto w
%     }{(\up T n) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{T' \evalsto \Nat b \qquad
    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
      \up{T\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             T %% {(\down {\Set[\infty]} T)}
             {(\down {T\,{\zero\infty}} t_z)}
             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
    }
\\[2ex]
  \ru{T' \evalsto \Nat b \qquad
    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
    }
\end{gather*}

@@

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{t \evalsto \up {\Nat \alpha} n \qquad
      \RNe k n m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{\alpha_v} \qquad
      \RSize k {\alpha_v} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{\alpha_v}u \qquad
      \RSize k {\alpha_v} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

\fbox{$\RSize k {\alpha_v} a$}
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k\dotminus(1+j)} + o}}
\end{gather*}

@@

A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive. The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A)$ the relation $\A$ is in fact an equivalence relation.

@@

\begin{lemma}[Closure properties of $\NE$ and $\NF$]
  \bla
  \begin{enumerate}
  \item $\lev k = \lev k \in \NE$.
  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' in \NE$.
%  \item If $d = d' \in \NF$
  \end{enumerate}
\end{lemma}

\[
\fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
  \msuchthat T \evalsto N \mand T' \evalsto N' \mand n = n' \in \NE \}
\]
\fbox{$\NAT(\alpha)$}
\begin{gather*}
  \ru{t \evalsto \zero \alpha \qquad
      t' \evalsto \zero{a'}
    }{t = t' \in \NAT(b+1)}
\qquad
  \ru{t \evalsto \suc \alpha u \qquad
      t' \evalsto \suc {a'} {u'} \qquad
      u = u' \in \NAT(b)
    }{t = t' \in \NAT(b+1)}
\\[2ex]
  \ru{t \evalsto \up{\Nat \alpha} n \qquad
      t' \evalsto \up{\Nat a'} n' \qquad
      n = n' \in \NE
    }{t = t' \in \NAT(b)}
\end{gather*}

@@

Let $\A$ be a PER and $\F$ a family of PERs over $\A$ such that
$\F(u) = \F(u')$ whenever $u = u' \in \A$.  We define
\[
  \fbox{$\PIAF$} \defas \{(t,t') \mid t\,u = t'\,u' \in \F(u) \mforall u = u' \in \A \}
  .
\]
This works also for $\A = \SIZE$.
For a family $\F$ over $\SIZE$ we also have the irrelevant function space
\[
  \fbox{$\FORALL\,\F$} \defas
  \{(t,t') \mid t \ann \alpha = t' \ann{a'} \in \F(b) \mforall \alpha,a',b \in \SIZE \}
  .
\]

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{T \evalsto N \qquad T' \evalsto N'
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All PERs involved here are closed under weak head expansion.

The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.  Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.

@@

Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its underlying simple type \cite{harperPfenning:equivalenceLF}. However, in the presence of universes and large eliminations, there is no underlying simple type.  Of course, we can take a type as its own shape, but we want at least that $\Nat \alpha$ and $\Nat b$ have the same shape even for different $\alpha, b$.  Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.

\fbox{$T \shape S$} for $T \in \SET[\ell]$.
\begin{gather*}
  \ru{T \evalsto \Nat \alpha \qquad
      S \evalsto \Nat b
    }{T \shape S}
\qquad
  \ru{T \evalsto N \qquad
      S \evalsto N'
    }{T \shape S}
\qquad
   \ru{T \evalsto \Set[\ell] \qquad
       S \evalsto \Set[\ell]
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      S \evalsto \piT{A'}{B'} \qquad
      A \shape A' \qquad
      B[u] \shape B'[u'] \mforall u = u' \in A
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \piT \Size B \qquad
      S \evalsto \piT{\Size}{B'} \qquad
      B[\alpha] \shape B'[\alpha] \mforall \alpha \in \SIZE
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \forallT B \qquad
      S \evalsto \forallT{B'} \qquad
      B[\alpha] \shape B'[a'] \mforall \alpha,a' \in \SIZE
    }{T \shape S}
\end{gather*}
% The last rule is the crucial one
\begin{lemma}[Reflexivity for shapes]
  If $T = T' \in \SET[\ell]$ then $T \shape T'$.
\end{lemma}

@@

For the main goal,
assume $\eta = \eta' \eeq \rho \in \Gamma$ and
show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
To this end, assume arbitrary $\alpha,a',b \in \SIZE$ and show
$t(\eta,\alpha) = t'(\eta',a') \in T(\rho,b)$.
Since $(\eta,\alpha) = (\eta',a') \eeq (\rho,b) \in \erext \Gamma \Size$,
we conclude by induction hypothesis.

>>>>

\fbox{$\Gamma \der t : T$}  (implies $\der \Gamma$ and $\resurrect \Gamma \der T$ [and $\resurrect \Gamma \der t : T$]. Note: no rule for $\Gamma \der \Size : s$.)
\begin{gather*}
  \ru{\der \Gamma
    }{\Gamma \der \Set[\ell] : \Set[\ell']
    }{\ell {<} \ell'}
\qquad
  \ru{\Gamma \der U : s \qquad
      \cext \Gamma U \der T : s
    }{\Gamma \der \piT U T : s}
\qquad
  \ru{\cext \Gamma \Size \der T : s
    }{\Gamma \der \epiT \Size T : s}
\qquad
  \ru{%\der \Gamma \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a : s}
\\[2ex]
  \rux{\der \Gamma \qquad \Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i : T
    }{T \not= \Size}
\qquad
  \ru{\eext \Gamma U \der t : T
    }{\Gamma \der \lambda t : \epiT U T}
\qquad
  \ru{\Gamma \der t : \piT U T \qquad
      \Gamma \der u : U
    }{\Gamma \der t\,u : T[u]}
% \qquad
%   \ru{\Gamma \der t : T \qquad
%       \Gamma \mid T \der e : T'
%     }{\Gamma \der t\,e : T'}
\qquad
  \ru{\Gamma \der t : \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a : T[a]}
\\[2ex] % \qquad
  \ru{\Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der \zero{a} : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der a : \Size \qquad
      \Gamma \der t : \Nat b
    }{\Gamma \der \suc a t : \Nat (b + 1)}
\qquad
  \ru{\Gamma \der t : \forallT T \qquad
      \Gamma \der a : \Size \qquad
      \resurrect\Gamma \der b : \Size
    }{\Gamma \der t \ann a : T[b]}
\\[2ex]
  \ru{\Gamma \der u : \Nat (a + 1) \qquad
      \Gamma \der T : \Nat (a + 1) \to \Set[\ell] \qquad
      \Gamma \der t_z : T \, (\zero{a}) \qquad
      \Gamma \der t_s : (t : \Nat a) \to T \, (\suc{a} t)
    }{\Gamma \der u\,\case[\ell] T {t_z} t_s : T \, u}
\\[2ex]
  \ru{\Gamma \der u : \Nat a \qquad
      \Gamma \der T : \FixK\,\ell \qquad %% \funT i \Size \Nat i \to \Set[\ell] \\
      \Gamma \der t : \FixT\, T
     }{\Gamma \der u\,\fix[\ell] T t : T \,a \, u}
\qquad
  \ru{\Gamma \der t : T \qquad
      \resurrect\Gamma \der T \leq T'
     }{\Gamma \der t : T'}
\end{gather*}
Note: Size is relevant in $T$ as we would like for instance $T\,i\,x = \Nat i$.
% \fbox{$\Gamma \mid U \der e : T$}
% \begin{gather*}
%   \ru{\cext \Gamma U \der T \qquad \Gamma \der u : U
%     }{\Gamma \mid \piT U T \der u : T[u]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \piT \Size T \der a : T[a]}
% \qquad
%   \ru{\cext \Gamma \Size \der T \qquad \Gamma \der a : \Size
%     }{\Gamma \mid \forallT T \der \ann a : T[a]}
% \\
%   \ru{
%     }{\Gamma \mid \Nat a \der \fix T t : T \ann a n}
% \end{gather*}

@@

  \ru{\Gamma \der t = t' : T \qquad
      \resurrect\Gamma \der T \leq T'
     }{\Gamma \der t = t' : T'}

@@

\fbox{$\Gamma \der T \leq T'$} (implies $\Gamma \der T,T'$)
% (implies $\resurrect \Gamma \der T,T'$)
\begin{gather*}
  \ru{\der \Gamma \qquad \ell {\leq} \ell'
    }{\Gamma \der \Set[\ell] \leq \Set[\ell']}
\qquad
  \ru{\Gamma \der a \leq b : \Size
    }{\Gamma \der \Nat a \leq \Nat b}
\qquad
  \ru{\Gamma \der T = T'
    }{\Gamma \der T \leq T'}
\\[2ex]
  \ru{\Gamma \der U' \leq U \qquad
      \cext \Gamma {U'} \der T \leq T'
    }{\Gamma \der \piT U T \leq \piT {U'} {T'}}
\qquad
  \ru{\cext \Gamma {\Size} \der T \leq T'
    }{\Gamma \der \epiT \Size T \leq \epiT {\Size} {T'}}
\qquad
  \ru{\Gamma \der T_1 \leq T_2 \qquad
      \Gamma \der T_2 \leq T_3
    }{\Gamma \der T_1 \leq T_3}
\qquad
\end{gather*}

@@

Weakenings \fbox{$\xi : \Gamma \leq \Delta$}  (implies $\der \Gamma$ and $\der \Delta$ and $\Gamma \der \xi : \Delta$).
\begin{gather*}
  \ru{\der \Gamma
    }{\sempty : \Gamma \leq \cempty}
\qquad
  \ru{\xi : \Gamma \leq \Delta \qquad
      \resurrect\Gamma \der T
    }{\xi \slift : \cext \Gamma T \leq \Delta}
\qquad
  \ru{\xi : \Gamma \leq \Delta \qquad
      \resurrect \Delta \der U \qquad
      \resurrect\Gamma \der T \leq U \xi
    }{(\xi \slift, \ind 0) : \cext \Gamma T \leq \cext \Delta U}
\end{gather*}

@@

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \alpha \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{t_z \evalsto w
    }{(\zero \alpha) @ \casett \evalsto w}
\qquad
  \ru{t_s\, t \evalsto w
    }{(\suc \alpha t) @ \casett \evalsto w}
\\[2ex]
  \ru{t\,\ann \alpha\,(\lambda x.\, x@ \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero \alpha, \suc \alpha u \}}
\end{gather*}
The rules added for NbE deal with elimination of delayed reflection:
\begin{gather*}
  \ru{T' \evalsto \piT U T
    }{(\up {T'} n) @ u \evalsto \up{T[u]} (n \, \down U u) }
\qquad
  \ru{T' \evalsto \piT \Size T
    }{(\up {T'} n) @ \alpha \evalsto \up{T[\alpha]} (n \, \alpha)}
\qquad
  \ru{T' \evalsto \forallT T
    }{(\up {T'} n) @ \ann \alpha \evalsto \up{T[\alpha]} (n \ann \alpha)}
%   \ru{T \evalsto \piT A B \qquad
%       \up{B[u]} (n \, \down A u) \evalsto w
%     }{(\up T n) @ u \evalsto w}
% \qquad
%   \ru{T \evalsto \piT \Size B \qquad
%       \up{B[u]} (n \, \alpha) \evalsto w
%     }{(\up T n) @ \alpha \evalsto w}
% \qquad
%   \ru{T \evalsto \forallT B \qquad
%       \up{B[\alpha]} (n \ann \alpha) \evalsto w
%     }{(\up T n) @ \ann \alpha \evalsto w}
\\[2ex]
  \ru{T' \evalsto \Nat \beta \qquad
    }{(\up {T'} n) @ \case[\ell] T {t_z} {t_s} \evalsto
      \up{T\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             T %% {(\down {\Set[\infty]} T)}
             {(\down {T\,{\zero\infty}} t_z)}
             {(\down {\funS x {\Nat \infty} T\,(\suc \infty x)} t_s)}
    }
\\[2ex]
  \ru{T' \evalsto \Nat \beta \qquad
    }{(\up {T'} n) @ \fix[\ell] T t \evalsto
      n\,\fix[\ell] {(\down {\FixK} T)} {(\down {\FixT\, T} t)}
    }
\end{gather*}

@@

\fbox{$\RNat k t v $}
\begin{gather*}
  \ru{t \evalsto \up {\Nat \alpha} n \qquad
      \RNe k n m
    }{\RNat k t m}
\qquad
  \ru{t \evalsto \zero{\alpha} \qquad
      \RSize k {\alpha} a
    }{\RNat k t {\zero a}}
\qquad
  \ru{t \evalsto \suc{\alpha}u \qquad
      \RSize k {\alpha} a \qquad
      \RNat k u v
    }{\RNat k t {\suc a v}}
\end{gather*}

\fbox{$\RSize k {\alpha} a$}
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k\dotminus(1+j)} + o}}
\end{gather*}

@@

A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive. The domain $\dom(\A)$ of the relation can be thought of the set of terms which define the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.  We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.

@@

\begin{lemma}[Closure properties of $\NE$ and $\NF$]
  \bla
  \begin{enumerate}
  \item $\lev k = \lev k \in \NE$.
  \item If $n = n' \in \NE$ and $d = d' \in \NF$ then $n\,d = n'\,d' \in \NE$.
%  \item If $d = d' \in \NF$
  \end{enumerate}
\end{lemma}

\[
\fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
  \msuchthat T \evalsto N \mand T' \evalsto N' \mand n = n' \in \NE \}
\]
\fbox{$\NAT(\alpha)$}
\begin{gather*}
  \ru{t \evalsto \zero \alpha \qquad
      t' \evalsto \zero{\alpha'}
    }{t = t' \in \NAT(\beta+1)}
\qquad
  \ru{t \evalsto \suc \alpha u \qquad
      t' \evalsto \suc {\alpha'} {u'} \qquad
      u = u' \in \NAT(\beta)
    }{t = t' \in \NAT(\beta+1)}
\\[2ex]
  \ru{t \evalsto \up{\Nat \alpha} n \qquad
      t' \evalsto \up{\Nat \alpha'} n' \qquad
      n = n' \in \NE
    }{t = t' \in \NAT(\beta)}
\end{gather*}

@@

Let $\A$ be a PER and $\F$ a family of PERs over $\A$ such that
$\F(u) = \F(u')$ whenever $u = u' \in \A$.  We define
\[
  \fbox{$\PIAF$} \defas \{(t,t') \mid t\,u = t'\,u' \in \F(u) \mforall u = u' \in \A \}
  .
\]
This works also for $\A = \SIZE$.
For a family $\F$ over $\SIZE$ we also have the irrelevant function space
\[
  \fbox{$\FORALL\,\F$} \defas
  \{(t,t') \mid t \ann \alpha = t' \ann{\alpha'} \in \F(\beta) \mforall \alpha,\alpha',\beta \in \SIZE \}
  .
\]

\begin{lemma}[PER Subsumption]
\label{lem:subsump} \hfill
\begin{enumerate}
\item If $\alpha \leq \beta$ then $\NAT(\alpha) \subseteq \NAT(\beta)$.
\item If $\F(\alpha) \leq \F'(\alpha)$ for all $\alpha \in \SIZE$, then $\FORALL \F \subseteq \FORALL \F'$.
\item IF $\A' \subseteq \A$ and $\F(u) \subseteq \F'(u)$ for all $u \in \A'$, then $\PIAF \subseteq \PI\,\A'\,\F'$.
\end{enumerate}
\end{lemma}

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{T \evalsto N \qquad T' \evalsto N'
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head expansion.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension is PER: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}

% The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.  Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
\begin{lemma}[Derivation independence of extension]
\label{lem:indep}
  If $\DD_1 :: T = T_1 \in \SET[\ell_1]$
  and $\DD_2 :: T_2 = T \in \SET[\ell_2]$ then $\EL[\ell_1](T) = \EL[\ell_2](T)$.
\end{lemma}
\begin{proof}
  By induction on $\DD_1$ and cases on $\DD_2$.
\end{proof}
Sine $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.

\begin{lemma}[Universe subsumption]
  \label{lem:usubsump}
  If $\ell \leq \ell'$ then $\SET[\ell] \subseteq \SET[\ell']$.
\end{lemma}
\begin{proof}
  Prove $T = T' \in \SET[\ell']$ by induction on $T = T' \in \SET[\ell]$.
  For the base types this is direct, let us look at the function space.
\[
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \qquad
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
\]
  By induction hypothesis, $A = A' \in \SET[\ell']$, and we have $\EL[\ell'](A) = \EL[\ell](A)$ by Lemma~\ref{lem:indep}.  Assuming $u = u' \in \EL(A)$, we get $B[u] = B'[u'] \in \SET[\ell']$ by induction hypothesis on $B[u] = B'[u'] \in \SET[\ell]$.
\end{proof}

@@

Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its underlying simple type \cite{harperPfenning:equivalenceLF}. However, in the presence of universes and large eliminations, there is no underlying simple type.  Of course, we can take a type as its own shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have the same shape even for different $\alpha, \beta$.  Thus, we define a relation $T \shape S$ to express that $S$ is a possible shape of type $T$.

\fbox{$T \shape S$} for $T \in \SET[\ell]$.
\begin{gather*}
  \ru{T \evalsto \Nat \alpha \qquad
      S \evalsto \Nat \beta
    }{T \shape S}
\qquad
  \ru{T \evalsto N \qquad
      S \evalsto N'
    }{T \shape S}
\qquad
   \ru{T \evalsto \Set[\ell] \qquad
       S \evalsto \Set[\ell]
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \piT A B \qquad
      S \evalsto \piT{A'}{B'} \qquad
      A \shape A' \qquad
      B[u] \shape B'[u'] \mforall u = u' \in A
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \piT \Size B \qquad
      S \evalsto \piT{\Size}{B'} \qquad
      B[\alpha] \shape B'[\alpha] \mforall \alpha \in \SIZE
    }{T \shape S}
\\[2ex]
  \ru{T \evalsto \forallT B \qquad
      S \evalsto \forallT{B'} \qquad
      B[\alpha] \shape B'[\alpha'] \mforall \alpha,\alpha' \in \SIZE
    }{T \shape S}
\end{gather*}
% The last rule is the crucial one
\begin{lemma}[Reflexivity for shapes]
  If $T = T' \in \SET[\ell]$ then $T \shape T'$.
\end{lemma}

@@

For the main goal,
assume $\eta = \eta' \eeq \rho \in \Gamma$ and
show $(\lambda t)\eta = (\lambda t')\eta' \in (\forall T)\rho$.
To this end, assume arbitrary $\alpha,\alpha',\beta \in \SIZE$ and show
$t(\eta,\alpha) = t'(\eta',\alpha') \in T(\rho,\beta)$.
Since $(\eta,\alpha) = (\eta',\alpha') \eeq (\rho,\beta) \in \erext \Gamma \Size$,
we conclude by induction hypothesis.
